---
title: "Assignment 5: Data Visualization"
author: "Joy Wu"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(here)
library(cowplot)
here()
PeterPaul.chem.nutrients <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
           stringsAsFactors = TRUE)

Niwot.Ridge.litter <- read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), 
           stringsAsFactors = TRUE)

#2 
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
Niwot.Ridge.litter$collectDate <- ymd(Niwot.Ridge.litter$collectDate)
class(PeterPaul.chem.nutrients$sampledate)
class(Niwot.Ridge.litter$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
mytheme <- theme_classic(base_size = 13) +
  theme(axis.text = element_text(color = "black"),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right")
theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

phosphorus.by.phosphate <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = po4, y = tp_ug, color = lakename, 
                                       shape = lakename, linetype = lakename)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point() +
  xlim(0, 75) +
  ylim(0, 125) + 
  labs(
    title = "Total Phosphorus vs. Phosphate",
    x = "Phosphate (PO4)",
    y = "Phosphorus (tp_ug)"
  )
print(phosphorus.by.phosphate)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

temp <- ggplot(PeterPaul.chem.nutrients,
               aes(x = factor(month, levels = 1:12, labels = month.abb),
                   y = temperature_C, color = lakename)) +
  geom_boxplot() +
  labs(title = "Temperature by Month", x = "Month", y = "Temperature (°C)") +
  theme(legend.position = "bottom")
print(temp)

tp <- ggplot(PeterPaul.chem.nutrients,
               aes(x = factor(month, levels = 1:12, labels = month.abb),
                   y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(title = "TP by Month", x = "Month", y = "Temperature (°C)") +
  theme(legend.position = "bottom")
print(tp)

tn <- ggplot(PeterPaul.chem.nutrients,
               aes(x = factor(month, levels = 1:12, labels = month.abb),
                   y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(title = "TN by Month", x = "Month", y = "Temperature (°C)") +
  theme(legend.position = "bottom")
print(tn)

plot_grid(
  temp + theme(legend.position = "none"),
  tp   + theme(legend.position = "none"),
  tn   + theme(legend.position = "bottom"),
  ncol = 1,
  align = "v",
  axis = "lr",
  rel_heights = c(1, 1, 1) 
)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Over seasons: Temperature increases in the summer, then decreases in the winter;
TP also increases in summer times, and TN shows a more steady trend throughout the seasons.
Between lakes: both lakes show a similar trend regarding temperature changes; Peter Lake 
generally has a higher TP and TN concentration than Paul Lake does over the months. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
niwot.subset <- Niwot.Ridge.litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(
    mapping = aes(
      x=collectDate,
      y=dryMass,
      color=nlcdClass
    )
  ) + 
  geom_point() +
  labs(
    title = "Niwot Ridge: Needle Litter Dry Mass Over Time",
    x = "Date",
    y = "Dry Mass",
    color = "NLCD class"
  ) +
  theme(legend.position = "bottom")
print(niwot.subset)


#7
niwot.facet <- Niwot.Ridge.litter %>%
  filter(functionalGroup == "Needles") %>%
  ggplot(
    mapping = aes(
      x=collectDate,
      y=dryMass
    )
  ) + 
  geom_point() +
  facet_wrap(facets = vars(nlcdClass)) + 
  labs(
    title = "Niwot Ridge: Needle Litter Dry Mass Over Time by NLCD Class",
    x = "Date",
    y = "Dry Mass"
  ) +
  theme(legend.position = "bottom")
print(niwot.facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 7 is more effective because it identifys patterns within each NLCD class and also allows for cross comparison between classes as they are placed next to each other. 