---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Joy Wu"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(lubridate)
options(scipen = 4)
library(here)
here()

chem.phys <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors = TRUE)
chem.phys$sampledate<- as.Date(chem.phys$sampledate, format = "%m/%d/%y")
class(chem.phys$sampledate)

#2
mytheme <- theme_classic(base_size = 13) +
  theme(axis.text = element_text(color = "black"),
        plot.background = element_rect(fill = "white", color = NA),
        legend.position = "right")
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes
Ha: Mean lake temperature recorded during July changes with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
chem.phys.month <- mutate(chem.phys, month = month(sampledate))

chem.phys.july <- chem.phys.month %>%
  filter(month == 7) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit()

#5
temp.depth <- ggplot(chem.phys.july, aes(x = depth, y = temperature_C)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  ylim(0, 35)+
  labs(title = "Temperature by depth",
    x = "Depth",
    y = "Temperature (°C)")
print(temp.depth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The general trend of the figure suggests that there is a negative relationship between temperature and depth. 
However, the distribution of points does not suggest a clear linearity of the trend. Stacked y value points for each x value points does not show clear linearity.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
temp.depth.regression <- lm(chem.phys.july$temperature_C ~
                              chem.phys.july$depth)
summary(temp.depth.regression)
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The results show that about 74% of the variability in temperature is explained by changes in depth. There are 9726 degrees of freedom of which this finding is based. The p-value (less than 0.05) indicates that the relationship between temperature and depth is statistically significant. Results show that for every 1 meter change in depth, the temperature will change by 1.95 degrees Celsius.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp.aic <- lm(data = chem.phys.july, temperature_C ~ depth + year4 + daynum)
step(temp.aic)

#10
temp.model <- lm(data = chem.phys.july, temperature_C ~ depth + year4 + daynum)
summary(temp.model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The results show that, the final set of explanatory variables include all three variables (daynum, depth, and year4) as that combination (not removing any variables) already gave us the lowest AIC.

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
temp.anova <- aov(data = chem.phys.july, temperature_C ~ lakename)
summary(temp.anova)

temp.anova2 <- lm(data = chem.phys.july, temperature_C ~ lakename)
summary(temp.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, we can see from the results of the linear model that the mean of different lakes are very different. 
Each lake has a different mean value from each other, as can  be seen from the "estimated std." column, 
where the differences in mean are shwon.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
temp.lake <- ggplot(chem.phys.july, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE)+
  ylim(0, 35)+
  labs(title = "Temperature by depth",
    x = "Depth",
    y = "Temperature (°C)") +
  theme(
    legend.position = "bottom")
print(temp.lake)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
tukey.lake.mean <- TukeyHSD(temp.anova)
tukey.lake.mean

peter.similar <- HSD.test(temp.anova, "lakename", group = TRUE)
peter.similar
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: From the group test, we can tell that Ward Lake and Paul Lake have the same mean
temperature, statistically speaking, as Peter Lake. We can tell that because these two lakes have the
same letter as Peter Lake in "groups".

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: I did a little bit research online and found that another test we can conduct to explore the difference between
mean temperature of Peter and Paul Lake is a paired t-test. It is commonly used to determine whether there is a 
significant difference between the means of two related or "paired" groups. This method works better for comparing 
Peter Lake and Paul Lake since they are in the same "group".



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
crampton.ward.july <- chem.phys.july %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")
crampton.ward.july

july.twosample <- t.test(crampton.ward.july$temperature_C ~ crampton.ward.july$lakename)
july.twosample
```

>Answer: The results indicate that these two lakes have different mean temperature in July. Crampton has 15.35, and Ward
has 14.46.
